---
title: "Finalproject_DS700_Nanduri_aditya"
author: "Aditya"
date: "4/29/2020"
output:
  word_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Read in dataset file

```{r}
library(readr)
library(dplyr)
library(mice)
library(VIM)
require(forecast)
```
***In Excel we have done cleaning part (removal of non numeric values and outliers) and Calculations mentioned in explanation document.***
*****In this R document we will impute missing values and create models and then compare them*****

```{r}
abbeville = (read_csv("Dataset-cleaneddata.csv"))
# what's in the data
summary(abbeville)
```
Before we start any furether analysis , lets do a quick plot to check for any outliers
```{r}

plot(abbeville[c(2,1)])

# 2008 shows an outlier but in document it was mentioned that HC received an unusually large n umber of requests since neighboring   New Orleans HC was closed due to hurricane.This will skew forecast accuracy. So, I am removing it and imputing this value as well.
```
```{r}
#removing 2008 oct data point
abbeville[which(abbeville[ ,2] == '2008' & abbevillenew[ ,3] == '10' ),1] <- NA
```

```{r}
# Lets get a better understanding of the pattern of missing data
md.pattern(abbeville)
# output of above statement tells us that 86 samples are complete and 11 have missing values that too in Infcoming examination column
aggr_plot <- aggr(abbeville, col=c('navyblue','red'), numbers=TRUE, sortVars=TRUE, labels=names(abbeville), cex.axis=.7, gap=3, ylab=c("Histogram of missing data","Pattern"))
# above statement gives better visualization about missing data that 89% of data is available
```

```{r}

# changing column name as imputation function wont work if there is space in column name
colnames(abbeville)[which(names(abbeville) == "Incoming Examinations")] <- "Incoming_Examinations"
#Imputing the missing data, using predictive mean matching imputation method
tempdata = mice(abbeville, m =5, method = 'pmm', maxit = 5, seed=500)

```

```{r}
summary(tempdata)
tempdata$imp$Incoming_Examinations
abbevillenew <- complete(tempdata,1)# make a new data set with cleaned and imputed values

```

```{r}
# calculating Dec 2009-Feb 2010 data
# Total number of cases from Dec 2009 - Feb 2010 is 5129
# Calculate what percentage of cases in Dec 2008 is sum of (Dec 2008,Jan 2009, Feb 2010) - W1
# Calculate what percentage of cases in Jan 2009 is sum of (Dec 2008,Jan 2009, Feb 2010) - W2
# Calculate what percentage of cases in Feb 2009 is sum of (Dec 2008,Jan 2009, Feb 2010) - W3
# Finally calculate Dec 2009 as W1 * 5129, similarly Jan 2010 as W2 * 5129 ans Feb 2010 as W3 * 5129
# with this approach I belive would preserve trend and seasonality


Dec2008.rows = which(abbevillenew[ ,2] == '2008' & abbevillenew[ ,3] == '12' )
Dec2008 = abbevillenew[Dec2008.rows, 1]
Jan2009.rows = which(abbevillenew[ ,2] == '2009' & abbevillenew[ ,3] == '1' )
Jan2009 = abbevillenew[Jan2009.rows, 1]
Feb2009.rows = which(abbevillenew[ ,2] == '2009' & abbevillenew[ ,3] == '2' )
Feb2009 = abbevillenew[Feb2009.rows, 1]
W1 = Dec2008 / sum(Dec2008 + Jan2009 + Feb2009 )
W2 = Jan2009 / sum(Dec2008 + Jan2009 + Feb2009 )
W3 = Feb2009 / sum(Dec2008 + Jan2009 + Feb2009 )

abbevillenew[which(abbevillenew[ ,2] == '2009' & abbevillenew[ ,3] == '12' ),1] <- round(W1 * 5129)
abbevillenew[which(abbevillenew[ ,2] == '2010' & abbevillenew[ ,3] == '1' ),1] <- round(W2 * 5129)
abbevillenew[which(abbevillenew[ ,2] == '2010' & abbevillenew[ ,3] == '2' ),1] <- round(W3 * 5129)
```

```{r}
write.csv(abbevillenew, "Abbeville_Cleansed_imputed.csv")
```

```{r}
finaldata = read.csv("Abbeville_Cleansed_imputed.csv")
plot(finaldata[ ,2], ylab = "Heart Exams", xlab = "Index", main ="Scatterplot of Heart Exams")
```

```{r}
# Build a time series from data and  view Trend and seasonality
finaldata_ts = ts(finaldata[ ,2], start=2006, frequency=12)
plot(finaldata_ts, ylab = "Heart Exams", xlab = "Years", main = "Time Series Plot of Heart Exams")

```

```{r}
# To decide on whch model to use we need to look at seasonality and trend seperately
plot(decompose(finaldata_ts))
# From the plot we understand that there is additive seasonality and additive trend ( upward linear trend with additive seasonality)
```

***when do we used which type of model***
***When there is no trend and no seasonality  we use Moving average, Simple exponential smoothing***
***When there is a trend and seasonality we use  arima(after looking for auto regression) and Holt's winter exponential smoothing***
```{r}
# Holt-Winters Expotential Smoothing with trending - Training model
finaldata_hw = HoltWinters(finaldata_ts, gamma = FALSE)

# Holt-Winters Forecasting for 2014 
finaldata_hw_forecast = forecast(finaldata_hw, h=12)
plot(finaldata_hw)

```

```{r}
plot(finaldata_hw_forecast, ylab = "Heart Exams", xlab = "Years")

```
 Holt-Winters model, including the 80% (dark gray) and 95% (light gray) confidence intervals (CI)
```{r}
summary(finaldata_hw_forecast)
#RSME : 313.4226
#MAPE : 15.90158
```

```{r}
# Correlation plots - not stationary - confirms trend
acf(finaldata_ts) # autoregression present, large spike at initial that decays towards 0 is autoregressive
pacf(finaldata_ts) # autoregressive and moving average, large spike in 1st 2 variables that decays to 0 is MA
```

```{r}
# Diffing to transform data to account for October 2008 spike in data
# This will fix time series with trends or drift
# Optimal number for diffing
ndiffs(finaldata_ts)

# Plot with diffing
plot(diff(finaldata_ts,1))
# optimal diffing is 1
```

```{r}
# Fit ARIMA model - let R decide best model
final.arima = auto.arima(finaldata_ts)
final.arima
#fitted model
fit = final.arima$fitted
# plot of fit
plot(finaldata_ts, ylab = "Heart Exams", xlab = "Years", main = "Time Series Plot of Heart Exams")
lines(fit, col='red')
```

```{r}
# Plot residuals in acf and pacf - look vs bounds
acf(final.arima$residuals)
pacf(final.arima$residuals)
```

```{r}
# ARIMA Forecast for next 12 months
hc.arima.forecast = forecast(final.arima, h=12)
plot(hc.arima.forecast,ylab = "Heart Exams", xlab = "Years")
summary(hc.arima.forecast)
#RSME : 303.3947
#MAPE : 15.12695
```
ARIMA generated forecast with 80% and 95% CI

```{r}

```
Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
